---
title: "FE 620 Final Project"
subtitle: "I pledge my honor that I have abided by the Stevens Honor System"
author: "Author: "
date: "Last compiled on `r format(Sys.time(), '%B %d, %Y')`"
output: html_document
runtime: shiny
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(moments)
library(plotly)
```

```{r import_data}
df = readxl::read_xlsx("/Users/andrew/Documents/SCHOOL/2023/Fall 2023/FE620/Final Project/grid1_lmr4qsjv.xlsx", skip = 1, col_names = T)
```

```{r}
head(df)
```
## Data Cleaning

```{r parse}
# Parse column names
colnames(df) = sapply(strsplit(colnames(df), "\\.{3}"), function(x) x[1])
# Remove null rows
df = df %>% subset(Strike != "ERROR(6)")
```

```{r split}
# Split dataset into calls and puts
calls = df %>%
  dplyr::select(1:(ncol(df) %/% 2))

puts = df %>%
  dplyr::select((1 + ncol(df) %/% 2):ncol(df))
```

```{r mid_price}
# Option price as mid price
calls = calls %>%
  dplyr::mutate(Price = (calls$Bid + calls$Ask) / 2, IVM = IVM / 100) %>%
  dplyr::select(-Bid, -Ask, -Last, -Volm)
puts = puts %>%
  dplyr::mutate(Price = (puts$Bid + puts$Ask) / 2, IVM = IVM / 100) %>%
  dplyr::select(-Bid, -Ask, -Last, -Volm)
```

```{r calc_tau}
# Tau in annual units from contract expiration
data_date = as.Date("12/01/23", "%m/%d/%y")
calls = calls %>%
  dplyr::mutate(tau = as.numeric(as.Date(sub(".*?(\\d{1,2}/\\d{1,2}/\\d{2,4}).*", "\\1", Ticker), "%m/%d/%y") - data_date) / 252) %>% dplyr::select(-Ticker) %>% dplyr::mutate_all(as.numeric)
puts = puts %>%
  dplyr::mutate(tau = as.numeric(as.Date(sub(".*?(\\d{1,2}/\\d{1,2}/\\d{2,4}).*", "\\1", Ticker), "%m/%d/%y") - data_date) / 252) %>% dplyr::select(-Ticker) %>% dplyr::mutate_all(as.numeric)
```

```{r}
max(calls$tau) * 12
```

## Pricing Model

Since the data we're working with does not have options of a time-to-maturity greater than 3 months, and all our data is from the option chain of one day, we can use a constant risk-free rate rather than a stochastic rate. As of December 01, 2023 (the day the data was pulled), the 1-month, 2-month, 3-month and 4-month treasury yields are $\sim 5.4%$, which is what we will use as our proxy for the risk-free rate in our models. The spot price was also \$458.65 at the time the data was pulled. Additionally, SPY pays dividends quarterly and proportional to the dividends paid by the companies in the index. For simplicity, I've set the dividend amount to the average historical amount over the past 4 quarters, multiplied in with the growth factor.

```{r init_params}
rf = 0.054
S0 = 458.65
div_freq = 0.25
div_amt = 1.61
N = 200
```

```{r}
h<-calls$tau/N
calls<-cbind(calls,h)
head(calls)
```

```{r option_pricing}
american_option_price = function(S0, r, sigma, tau, K, steps, opt_type, greeks = TRUE,div_sched = NULL) {
  
  # Validating the input parameters
  if (S0 <= 0 || r <= 0 || sigma <= 0 || tau <= 0 || K <= 0 || steps <= 0) {
    stop("Input parameters should be positive and non-zero.")
  }
  
  # Calculating the necessary variables
  dt = tau / steps  # Time step size
  u = exp(sigma * sqrt(dt))  # Up factor
  d = 1 / u  # Down factor
  p = (exp(r * dt) - d) / (u - d)  # Probability of up movement
  
  # Adjust stock prices for dividends
  if (!is.null(div_sched)) {
    for (i in seq_along(div_sched$time)) {
      time = div_sched$time[i]
      amount = div_sched$amount[i]
      discount_factor = exp(-r * time)
      S0 = S0 - amount * discount_factor
    }
  }
  
  # Creating a 2D array to store the stock prices at each node of the tree
  stock_prices <<- matrix(0, nrow = steps + 1, ncol = steps + 1)
  
  # Calculating the stock prices at each node of the tree
  for (i in seq(1, steps + 1)) {
    for (j in seq(1, i)) {
      stock_prices[i, j] <<- S0 * (u ^ (i - j)) * (d ^ (j - 1))
    }
  }
  
  # Creating a 2D array to store the option values at each node of the tree
  option_values <<- matrix(0, nrow = steps + 1, ncol = steps + 1)
  
  # Calculating the option values at each node of the tree
  for (i in seq(steps, 0, -1)) {
    for (j in seq(1, i + 1)) {
      intr_val = ifelse(opt_type == "put", K - stock_prices[i, j], stock_prices[i, j] - K)
      if (i == steps) {
        # At the last step, the option value is the maximum of 0 and the intrinsic value
        option_values[i, j] <<- max(0, intr_val)
      } else {
        # For other steps, the option value is the maximum of the intrinsic value and the discounted expected value
        option_values[i, j] <<- max(intr_val,
                                   (p * option_values[i + 1, j] + (1 - p) * option_values[i + 1, j + 1]) * exp(-r * dt)
        )
      }
    }
  }
  if(greeks == TRUE){
    # Returning the option value at the root node of the tree
    fuu<-option_values[3,1]
    fud<-option_values[3,2]
    fdd<-option_values[3,3]
    Suu<-stock_prices[3,1]
    Sud<-stock_prices[3,2]
    Sdd<-stock_prices[3,3]
    
    t1<-(fuu-fud)/(Suu-Sud)
    t2<-(fud-fdd)/(Sud-Sdd)
    t3<-(Suu-Sdd)/2
    
    return(data.frame(est = option_values[1,1], Delta =(option_values[2,1]-option_values[2,2])/(stock_prices[2,1]-stock_prices[2,2]) , Gamma = (t1-t2)/t3, Theta = (fud-option_values[1,1])/(2*dt)))
  }
  else{
    return(option_values[1,1])
  }
}
```


```{r create_div_schedule}
div_sched_calc = function(tau, freq, amt, time_till_next_exdiv=freq) {
  if (tau < time_till_next_exdiv) return(NULL)
  time = seq(time_till_next_exdiv, tau, freq)
  amount = rep(amt, length(time))
  return(list(time = time, amount = amount))
}
```

```{r apply_call_pricing}
output<-dplyr::bind_rows(apply(calls, 1, function(row)
  american_option_price(
    S0 = S0,
    r = rf,
    sigma = row["IVM"],
    tau = row["tau"],
    K = row["Strike"],
    steps = N,
    opt_type = "call",
    greeks = TRUE,
    # data was from 12/01/23, next ex_div data is 12/16/23 -> 16-1=15 days
    div_sched = div_sched_calc(row["tau"], div_freq, div_amt, 11/252)
  )
))

calls<-cbind(calls,output)
```

```{r apply_put_pricing}
puts$est <- apply(puts, 1, function(row)
  american_option_price(
    S0 = S0,
    r = rf,
    sigma = row["IVM"],
    tau = row["tau"],
    K = row["Strike"],
    steps = N,
    opt_type = "put",
    greeks = FALSE,
    # data was from 12/01/23, next ex_div data is 12/16/23 -> 16-1=15 days
    div_sched = div_sched_calc(row["tau"], div_freq, div_amt, 11/252)
  )
)

```

## Calculating the Greeks

We calculated delta, gamma, and theta using analytical approximations which can be seen below. This was chosen in order to save on computation time, as to numerically compute each greek takes roughly double the time of computing the price of the option itself, and these formulas give good approximations.

$$\Delta = \frac{f_u - f_d}{S_u - S_d}$$

$$\Gamma = \frac{\frac{f_{uu}-f_{ud}}{S_{uu}-S_{ud}}-\frac{f_{ud}-f_{dd}}{S_{ud}-S_{dd}}}{\frac{S_{uu}-S_{dd}}{2}}$$

$$\Theta = \frac{f_{ud}-f}{2 \cdot \frac{\tau}{n}}$$

 We calculated vega and rho using finite difference methods. The formula for these are as follows...

$$\upsilon = \frac{A(s,\sigma +\Delta \sigma)-A(s, \sigma - \Delta \sigma)}{2 \cdot \Delta \sigma}$$

$$\rho = \frac{A(s,r +\Delta r)-A(s, r - \Delta r)}{2 \cdot \Delta r}$$

```{r greeks, echo=FALSE}

#Calculate Vega
vega<-(apply(calls, 1, function(row)
  american_option_price(
    S0 = S0,
    r = rf,
    sigma = row["IVM"]+(row["IVM"]*row["h"]),
    tau = row["tau"],
    K = row["Strike"],
    steps = N,
    opt_type = "call",
    greeks = FALSE,
    # data was from 12/01/23, next ex_div data is 12/16/23 -> 16-1=15 days
    div_sched = div_sched_calc(row["tau"], div_freq, div_amt, 11/252)
  )
)-apply(calls, 1, function(row)
  american_option_price(
    S0 = S0,
    r = rf,
    sigma = row["IVM"]-(row["IVM"]*row["h"]),
    tau = row["tau"],
    K = row["Strike"],
    steps = N,
    opt_type = "call",
    greeks = FALSE,
    # data was from 12/01/23, next ex_div data is 12/16/23 -> 16-1=15 days
    div_sched = div_sched_calc(row["tau"], div_freq, div_amt, 11/252)
  )
))/(2*calls$IVM*calls$h)

#Calculate Rho
rho<-(apply(calls, 1, function(row)
  american_option_price(
    S0 = S0,
    r = rf+(rf*row["h"]),
    sigma = row["IVM"],
    tau = row["tau"],
    K = row["Strike"],
    steps = N,
    opt_type = "call",
    greeks = FALSE,
    # data was from 12/01/23, next ex_div data is 12/16/23 -> 16-1=15 days
    div_sched = div_sched_calc(row["tau"], div_freq, div_amt, 11/252)
  )
)-apply(calls, 1, function(row)
  american_option_price(
    S0 = S0,
    r = rf-(rf*row["h"]),
    sigma = row["IVM"],
    tau = row["tau"],
    K = row["Strike"],
    steps = N,
    opt_type = "call",
    greeks = FALSE,
    # data was from 12/01/23, next ex_div data is 12/16/23 -> 16-1=15 days
    div_sched = div_sched_calc(row["tau"], div_freq, div_amt, 11/252)
  )
))/(2*rf*calls$h)

calls<-cbind(calls,vega,rho)
```

```{r}
head(calls)
```

## Evaluating Model Output

### Market vs Model

Looking at the below plot of our model prices vs observed market prices, we see that out model provides a good estimation, but this estimation begins to vary as tau increases. 

```{r}
plot(calls$Price, col="blue",ylim = c(0,20),main = "Market vs Estimation(Calls)",xlab = "Tau", ylab = "Price in $")
points(calls$est, col="red")
legend(0, 20, legend=c("Market Price", "Estimation"),  
       fill = c("blue","red") 
)
```

```{r}
plot(puts$Price, col="blue",ylim = c(0,20),main = "Market vs Estimation(Puts)",xlab = "Tau", ylab = "Price in $")
points(puts$est, col="red")
legend(0, 20, legend=c("Market Price", "Estimation"),  
       fill = c("blue","red") 
)
```

To better identify and quantify this variation, we can plot the MSE of each estimation, which are separated by tau(time to expiry).

```{r}
diff_squared<-(calls$Price-calls$est)^2
MSE<-c(sum(diff_squared[1:20])/20,sum(diff_squared[21:40])/20,sum(diff_squared[41:60])/20,sum(diff_squared[61:80])/20)
tau<-unique(calls$tau)

plot(tau,MSE, main = "MSE vs Tau(Calls)")
```

Here we can see that our model prices the option with only a two week expiration(the smallest $\tau$) the best, with an average difference of only a couple cents. As tau increases, we can see our model become less and less precise, pricing the longest-dated option dollars away from the true market value.

```{r}
diff_squared2<-(puts$Price-puts$est)^2
MSE2<-c(sum(diff_squared2[1:20])/20,sum(diff_squared2[21:40])/20,sum(diff_squared2[41:60])/20,sum(diff_squared2[61:80])/20)
tau2<-unique(puts$tau)

plot(tau2,MSE2, main = "MSE vs Tau(Puts)")
```

The same trend can be seen when pricing puts, though to less of a degree as tau increases. This increase in error can likely be attributed to our use of constant interest rates and volatility, which are exacerbated when there is more time for the model to account for. It's also important to note that for both call and put options, we consistently overprice and never under price. This could be attributed toward high volatility or rates the day we took data, with one or both expected to tame in the near future, but this deserves further investigation.

## Evaluating the Greeks

### Delta
```{r}
plot(calls$Strike[1:20],calls$Delta[1:20], main = "Delta vs Strike",col = "lightblue",xlab = "Strike", ylab = "Delta")
points(calls$Strike[21:40],calls$Delta[21:40],col = "blue")
points(calls$Strike[41:60],calls$Delta[41:60],col = "darkblue")
points(calls$Strike[61:80],calls$Delta[61:80],col = "black")
legend(450, .4, legend=c("Tau 1", "Tau 2", "Tau 3", "Tau 4"),  
       fill = c("lightblue","blue","darkblue","black") 
)
```

Looking at the plot of delta vs strike, we can see the delta of the options for each expiration date decrease the further out of the money the option is. 

### Gamma
```{r}
plot(calls$Strike[1:20],calls$Gamma[1:20],ylim=c(0,.04), main = "Gamma vs Strike",col = "chocolate1",xlab = "Strike", ylab = "Gamma")
points(calls$Strike[21:40],calls$Gamma[21:40],col = "chocolate2")
points(calls$Strike[41:60],calls$Gamma[41:60],col = "chocolate3")
points(calls$Strike[61:80],calls$Gamma[61:80],col = "chocolate4")
legend(450, .01, legend=c("Tau 1", "Tau 2", "Tau 3", "Tau 4"),  
       fill = c("chocolate1","chocolate2","chocolate3","chocolate4") 
)
```

### Theta
```{r}
plot(calls$Strike[1:20],calls$Theta[1:20],ylim = c(-62,-30), main = "Theta vs Strike",col = "coral1",xlab = "Strike", ylab = "Theta")
points(calls$Strike[21:40],calls$Theta[21:40],col = "coral2")
points(calls$Strike[41:60],calls$Theta[41:60],col = "coral3")
points(calls$Strike[61:80],calls$Theta[61:80],col = "coral4")
legend(465, -50, legend=c("Tau 1", "Tau 2", "Tau 3", "Tau 4"),  
       fill = c("coral1","coral2","coral3","coral4") 
)
```


### Vega
```{r}
plot(calls$Strike[1:20],calls$vega[1:20],ylim=c(30,95), main = "Vega vs Strike",col = "burlywood",xlab = "Strike", ylab = "Vega")
points(calls$Strike[21:40],calls$vega[21:40],col = "burlywood2")
points(calls$Strike[41:60],calls$vega[41:60],col = "burlywood3")
points(calls$Strike[61:80],calls$vega[61:80],col = "burlywood4")
legend(449, 99, legend=c("Tau 1", "Tau 2", "Tau 3", "Tau 4"),  
       fill = c("burlywood","burlywood2","burlywood3","burlywood4") 
)
```

### IVM
```{r}
plot(calls$Strike[1:20],calls$IVM[1:20],ylim=c(.1,.14), main = "IV vs Strike",col = "aquamarine1",xlab = "Strike", ylab = "IV")
points(calls$Strike[21:40],calls$IVM[21:40],col = "aquamarine2")
points(calls$Strike[41:60],calls$IVM[41:60],col = "aquamarine3")
points(calls$Strike[61:80],calls$IVM[61:80],col = "aquamarine4")
legend(465, .14, legend=c("Tau 1", "Tau 2", "Tau 3", "Tau 4"),  
       fill = c("aquamarine1","aquamarine2","aquamarine3","aquamarine4") 
)
```

```{r, eval=FALSE}
Moneyness <- unique((S0-calls$Strike)/S0)
Maturity <- unique(calls$tau*252)
IV <- as.matrix(data.frame(calls$IVM[1:20],calls$IVM[21:40],calls$IVM[41:60],calls$IVM[61:80]))



axx = list(title = "x: Moneyness (in %, Neg. values are OTM call)")
axy = list(title = "y: Time to Maturity (in days)")
axz = list(title = "z: Implied Volatility")

plot_ly(x=~Moneyness, y=~Maturity, z=~IV) %>% 
    add_surface() %>%
    layout(title = "December 1st, 2023 SPY Option Vol Surface",
           scene = list(xaxis=axx,yaxis=axy,zaxis=axz))

```

### Rho
```{r}
plot(calls$Strike[1:20],calls$rho[1:20],ylim=c(0,80), main = "Rho vs Strike",col = "cyan1",xlab = "Strike", ylab = "Rho")
points(calls$Strike[21:40],calls$rho[21:40],col = "cyan2")
points(calls$Strike[41:60],calls$rho[41:60],col = "cyan3")
points(calls$Strike[61:80],calls$rho[61:80],col = "cyan4")
legend(465, 80, legend=c("Tau 1", "Tau 2", "Tau 3", "Tau 4"),  
       fill = c("cyan1","cyan2","cyan3","cyan4") 
)
```

## Delta Hedging

```{r, echo = FALSE}
getSymbols("SPY", from = "2023-12-01")
S<-SPY$SPY.Adjusted

#Choose K = 458 for this exercise
tau1_K458_puts<-c(puts[9,c(1:4)])

delta_hedge<-american_option_price(
    S0 = S$SPY.Adjusted[1],
    r = rf,
    sigma = tau1_K458_calls$IVM,
    tau = tau1_K458_calls$tau-(1/252),
    K = tau1_K458_calls$Strike,
    steps = N,
    opt_type = "put",
    greeks = TRUE,
    # data was from 12/01/23, next ex_div data is 12/16/23 -> 16-1=15 days
    div_sched = div_sched_calc(tau1_K458_calls$tau, div_freq, div_amt, 11/252)
  )

for(i in c(2:length(S$SPY.Adjusted))){
  delta_hedge<-rbind(delta_hedge,american_option_price(
    S0 = S$SPY.Adjusted[i],
    r = rf,
    sigma = tau1_K458_calls$IVM,
    tau = tau1_K458_calls$tau-(i/252),
    K = tau1_K458_calls$Strike,
    steps = N,
    opt_type = "put",
    greeks = TRUE,
    # data was from 12/01/23, next ex_div data is 12/16/23 -> 16-1=15 days
    div_sched = div_sched_calc(tau1_K458_calls$tau-(i/252), div_freq, div_amt, 11/252)
  ))
  
}

Put_diff<-data.frame(c(0,delta_hedge[-1,1]-delta_hedge[-9,1]))
delta_1<-data.frame(c(0,delta_hedge[-9,2]))
price_diff<-data.frame(c(0,as.vector(S[-1])-as.vector(S[-9])))

delta_hedge<-cbind(data.frame(S),delta_hedge[,c(1,2)],Put_diff,delta_1*price_diff)

colnames(delta_hedge)<-c("S", "Est", "Delta", "P(t) - P(t-1)", "Pi(t) - Pi(t-1)")
                                                               
```
